Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-11-12 paper/paper.tex
recompile paper to make sure its current

\title{Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict\thanks{John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at \url{https://github.com/andybega/Blair-Sambanis-replication}.}}
\author{Andreas Beger\footnote{Predictive Heuristics, \href{mailto:adbeger@gmail.com}{\nolinkurl{adbeger@gmail.com}}.} \and Richard K. Morgan\footnote{Varieties of Democracy Institute, University of Gothenburg, \href{mailto:rick.morgan2@gmail.com}{\nolinkurl{rick.morgan2@gmail.com}}.} \and Michael D. Ward\footnote{Predictive Heuristics, Duke University, University of Washington, \href{mailto:michael.don.ward@gmail.com}{\nolinkurl{michael.don.ward@gmail.com}}. Corresponding author.}}
\date{10 November 2020}

@@

Blair and Sambanis (n.d., hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed, they assert that with such theory forecasting is more accurate than has previously been possible. We re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives that they examine. It does worse. The reason for this is that they have made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkably accurately''. Indeed, B\&S argue that this theoretical model outperforms four other ``more mechanical'' alternatives. They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{PITF (e.g., Goldstone et al. 2010; Beger and Ward 2017), the W-ICEWS project (Ward et al. 2013), VIEWS (Hegre et al. 2019), and others, e.g., Beger, Morgan, and Maxwell (2020).} B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

As a result: (1) in the test set, all countries with names that alphabetically start after ``Tongo'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical. For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions weighted by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.

@@

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling (Chiba, Metternich, and Ward 2015; Beger et al. 2017). Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees. Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (\texttt{ntree}), the new random forest model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split. The split-population PITF random forest model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B\&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.

@@

Table \ref{tab:table1-nosmooth} (B\&S Table 1) shows the comparison of the Escalation model to other alternative models based on ICEWS event data indicators. The AUC-ROC values are based on original, non-smoothed ROC curves. Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}). The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B\&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves.

@@

\caption{\label{tab:table2-N}Number of valid test predictions for the escalation and structural alterantives models (B\&S Table 2)}
\centering
\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
 & Horizon & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Original model-specific cases}}\\
\hspace{1em} & 1 month & 13748 & 13155 & 13461 & 13748 & 13510\\
\cmidrule{2-7}
\hspace{1em} & 6 months & 2366 & 2264 & 2317 & 2366 & 2265\\
\cmidrule{1-7}
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Cases adjusted to common subset}}\\
\hspace{1em} & 1 month & 13062 & 13062 & 13062 & 13062 & 13062\\
\cmidrule{2-7}
\hspace{1em} & 6 months & 2250 & 2250 & 2250 & 2250 & 2250\\
\bottomrule
\end{tabular}
\end{table}

@@

B\&S interpret their results as follows (page 20) and we comment on the conclusions seratim:\footnote{We list the ``Overall, \ldots{}'' interpretation out-of-order for clarity.}

@@

For reference for the following discussion, we reproduce confusion matrices for the 2016-H1 forecasts in Table \ref{tab:table4}, but distinguish between civil war onsets (\(Y=1\)), non-events (\(Y=0\)) that correspond to countries with an ongoing civil war, and non-events (\(Y=0\)) that correspond to countries without a civil war, but which could potentially experience a civil war onset. The original B\&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group. Essentially all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are ongoing civil war cases which by definition \emph{cannot} experience another civil war onset. The alternative outcome coding B\&S evaluate, under ``Assuming Change'', produces the only two civil war onsets we would have observed in 2016-H1.

\begin{table}
\caption{Replication of B\&S Table 4: Confusion Matrices for Six-month Escalation Model forecasts for the first half of 2016.}\label{tab:table4}
\centering
\begin{tabular}{ll|c|c|l}
\multicolumn{1}{l}{\multirow{5}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$, no civil war & 134 & 17 \\
\cline{3-4}
& $Y=0$, ongoing & 2 & 13 \\
\cline{3-4}
& $Y=1$, onset & 0 & 0 \\
\cline{3-4}
% fixed, assuming change
\multicolumn{1}{l}{\multirow{5}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$, no civil war & 134 & 16 \\
\cline{3-4}
& $Y=0$, ongoing & 2 & 12 \\
\cline{3-4}
& $Y=1$, onset & 0 & 2 \\
\cline{3-4}
\end{tabular}
\end{table}

@@

\begin{quote}
Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries---even those with ongoing conflicts as of December 31, 2015--are included in the 2016 test set. As a result, 14 of the 30 countries on the list (and nine of the top 10) are cases of ongoing conflict.
\end{quote}

@@

Does it make sense to count a high probability for the risk of onset for a case with an ongoing civil war as a partial or complete ``win'' for the model? When creating the confusion matrices, B\&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events (``1''). In the text, B\&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases. This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g.~two separatists conflicts in different portions of the country. But this is not the case; ongoing conflicts in all other analyses are consistently dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.

@@

Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for actually predicting those outcomes. It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al. 2010), irregular leadership changes (Beger et al. 2017), and coups (Beger and Ward 2020).

Aside from the selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility regarding functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are \emph{the} correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B\&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory (see also Fariss and Jones 2018)?

@@

We encountered several issues in the code underlying the B\&S analysis. The problems we encountered are not subjective modeling choices. They bear no implications on the question of whether theory-based model are preferable. However, when we fix these issues and perform an updated analysis, the B\&S conclusions are all essentially overturned. In other words, B\&S findings are based on a faulty analysis, invalid, and should be disregarded as evidence in favor of theory-based models.

@@

See \texttt{1mo\_run\_escalation\_weighted\_PITF.R} line 4, where the PITF predictions are taken from the training data set (\texttt{train\$pred\_prob\_plus1}). The next line is a hack extending the shorter \texttt{weight} vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly in the 6-month version of this file.

@@

The intention is to train separate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest. Actually both random forests are trained on identical data, essentially production a random forest that just has N\_trees * 2.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Throughout the R code for the main analyses, the models predict a 1-time period lead version of the civil war onset variables (ending in "\_plus1``). This means that, for example, in the 6-month version of the data, a value of''1" for the "\_plus1" outcome variable for 2015-H2 actually refers to a civil war onset in the first half of 2016. One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV.
\item
  The nominally 2016-H1 forecasts are created at the bottom of \texttt{+master.R} using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2.
\item
  The scoring of the forecasts, whatever time period they are actually for, is done in a separate Stata do file, \texttt{6mo\_make\_confusion\_matrix.do}. Here, the scoring is done using the non-lead version of the DV. Thus there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
\end{enumerate}

The relevant variables in the data are ``incidence\_civil\_ns'' and ``incidence\_civil\_ns\_plus1'', which appears to be a 1-period lead (ie., \(t_+1\)) version of the dependent variable that is used in the actual prediction models. The incidence dependent variable contains both 0/1 and missing values. By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing. This reflects common practice.

>>>>

\title{Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict\thanks{John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at \url{https://github.com/andybega/Blair-Sambanis-replication}.}}
\author{Andreas Beger\footnote{Predictive Heuristics, \href{mailto:adbeger@gmail.com}{\nolinkurl{adbeger@gmail.com}}.} \and Richard K. Morgan\footnote{Independent Researcher, \href{mailto:rick.morgan2@gmail.com}{\nolinkurl{rick.morgan2@gmail.com}}.} \and Michael D. Ward\footnote{Predictive Heuristics, Duke University, University of Washington, \href{mailto:michael.don.ward@gmail.com}{\nolinkurl{michael.don.ward@gmail.com}}. Corresponding author.}}
\date{12 November 2020}

@@

Blair and Sambanis (n.d., hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed, they assert that with such theory forecasting is more accurate than has previously been possible. We re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives they examine. It does worse. The reason for this is that they made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to argue that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkably accurately''. Indeed, B\&S argue that this theoretical model outperforms four other ``more mechanical'' alternatives. They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{PITF (e.g., Goldstone et al. 2010; Beger and Ward 2017), the W-ICEWS project (Ward et al. 2013), ViEWS (Hegre et al. 2019), and others, e.g., Beger, Morgan, and Maxwell (2020).} B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

As a result: (1) in the test set, all countries with names that alphabetically start after ``Tonga'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical. For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions weighted by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.

@@

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling (Chiba, Metternich, and Ward 2015; Beger et al. 2017). Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees. Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (\texttt{ntree}), the new random forest model after combination has 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split. The split-population PITF random forest model is, practically speaking, simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra (odd) randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

Thus, a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B\&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.

@@

Table \ref{tab:table1-nosmooth} (B\&S Table 1) shows the comparison of the Escalation model to other alternative models based on ICEWS event data indicators. The AUC-ROC values are based on original, non-smoothed ROC curves. Both the Average and the CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}). The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B\&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves.

@@

\caption{\label{tab:table2-N}Number of valid test predictions for the escalation and structural alternatives models (B\&S Table 2)}
\centering
\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
 & Horizon & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Original model-specific cases}}\\
\hspace{1em} & 1 month & 13748 & 13155 & 13461 & 13748 & 13510\\
\cmidrule{2-7}
\hspace{1em} & 6 months & 2366 & 2264 & 2317 & 2366 & 2265\\
\cmidrule{1-7}
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Cases adjusted to common subset}}\\
\hspace{1em} & 1 month & 13062 & 13062 & 13062 & 13062 & 13062\\
\cmidrule{2-7}
\hspace{1em} & 6 months & 2250 & 2250 & 2250 & 2250 & 2250\\
\bottomrule
\end{tabular}
\end{table}

@@

B\&S interpret their results as follows (page 20) and we comment on the conclusions seriatim:\footnote{We list the ``Overall, \ldots{}'' interpretation out-of-order for clarity.}

@@

For reference for the following discussion, we reproduce confusion matrices for the 2016-H1 forecasts in Table \ref{tab:table4}, but distinguish between civil war onsets (\(Y=1\)), non-events (\(Y=0\)) that correspond to countries with an ongoing civil war, and non-events (\(Y=0\)) that correspond to countries without a civil war, but which could potentially experience a civil war onset. The original B\&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group. Essentially, all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are ongoing civil war cases which by definition \emph{cannot} experience another civil war onset. The alternative outcome coding B\&S evaluate, under ``Assuming Change'', produces the only two civil war onsets we would have observed in 2016-H1.

\begin{table}
\caption{Replication of B\&S Table 4: Confusion Matrices for Six-month Escalation Model forecasts for the first half of 2016.}\label{tab:table4}
\centering
\begin{tabular}{ll|c|c|l}
\multicolumn{1}{l}{\multirow{5}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$, no civil war & 134 & 17 \\
\cline{3-4}
& $Y=0$, ongoing & 2 & 13 \\
\cline{3-4}
& $Y=1$, onset & 0 & 0 \\
\cline{3-4}
% fixed, assuming change
\multicolumn{1}{l}{\multirow{5}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$, no civil war & 134 & 16 \\
\cline{3-4}
& $Y=0$, ongoing & 2 & 12 \\
\cline{3-4}
& $Y=1$, onset & 0 & 2 \\
\cline{3-4}
\\
\end{tabular}
\end{table}

@@

\begin{quote}
Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries---even those with ongoing conflicts as of December 31, 2015---are included in the 2016 test set. As a result, 14 of the 30 countries on the list (and nine of the top 10) are cases of ongoing conflict.
\end{quote}

@@

Does it make sense to count a high probability for the risk of onset for a case with an ongoing civil war as a partial or complete ``win'' for the model? When creating the confusion matrices, B\&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events (``1''). In the text, B\&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases. This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g., two separatists' conflicts in different portions of the country. But this is not the case; ongoing conflicts in all other analyses are consistently dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.

@@

Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that are good predictors do not originate from prior theory. For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for actually predicting those outcomes. It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g., the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al. 2010), irregular leadership changes (Beger et al. 2017), and coups (Beger and Ward 2020).

Aside from the selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility regarding functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are \emph{the} correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and, to some extent, are capable of learning non-linear and interactive relationships. B\&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory (see also Fariss and Jones 2018)?

@@

We encountered several issues in the code underlying the B\&S analysis. The problems we encountered are not subjective modeling choices. They bear no implications on the question of whether theory-based models are preferable. However, when we fix these issues and perform an updated analysis, the B\&S conclusions are all essentially overturned. In other words, B\&S findings are based on a faulty analysis, invalid, and should be disregarded as evidence in favor of theory-based models.

@@

See \texttt{1mo\_run\_escalation\_weighted\_PITF.R} line 4, where the PITF predictions are taken from the training data set (\texttt{train\$pred\_prob\_plus1}). The next line is a hack extending the shorter \texttt{weight} vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions -- similarly in the 6-month version of this file.

@@

The intention is to train separate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest. Actually, both random forests are trained on identical data, essentially production a random forest that just has N\_trees * 2.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Throughout the R code for the main analyses, the models predict a 1-time period lead version of the civil war onset variables (ending in "\_plus1``). This means that, for example, in the 6-month version of the data, a value of''1" for the "\_plus1" outcome variable for 2015-H2 actually refers to a civil war onset in the first half of 2016. One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV.
\item
  The nominally 2016-H1 forecasts are created at the bottom of \texttt{+master.R} using the data for 2016-H1 and still the lead DV. Thus, the lead DV here refers to outcomes in 2016-H2.
\item
  The scoring of the forecasts, whatever time period they are actually for, is done in a separate Stata do file, \texttt{6mo\_make\_confusion\_matrix.do}. Here, the scoring is done using the non-lead version of the DV. Thus, there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
\end{enumerate}

The relevant variables in the data are ``incidence\_civil\_ns'' and ``incidence\_civil\_ns\_plus1'', which appears to be a 1-period lead (i.e., \(t_{+1}\)) version of the dependent variable that is used in the actual prediction models. The incidence dependent variable contains both 0/1 and missing values. By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing. This reflects common practice.
